Type Checker Generation using Reference Attribute Grammars

### 2024-01-16 MSc.tex
Further improvements

  This report contains a study on the usability of JastAdd to programmatically generate executable type checkers from a subset of possible typing rules.
  This aims to provide integration with existing compilers implemented with JastAdd, as well as a playground for type systems usable as an educational tool.

  % 'Syntax-directed' needs explaining if used, "a subset known as syntax-directed"
  We find RAGs to be a flexible model for this subset of typing rules,
  % This is too specific, works with small languages from research literature
  and our software was able to support simple values, nodes with premises placing demands on their children, and those with rules requiring type variables.
  It proved difficult to provide compile-time verification of the typing rules, and further work is required to extend the compiler to support type environments, a necessity for most real-world type systems.
}

@@

\section{Abstract syntax trees}\label{abstractsyntaxtrees}
\CR{What's a syntax?}
An abstract syntax tree (AST) is a tree-based representation of the syntax of a program.
Each node in the tree represents a syntactic construct of the language, such as expressions, statements or declarations and the branches represent the structure.
It is abstract in that it neglects to include all the concrete details of the syntax, but rather focuses on the structure of the program.
Structural elements of the source code, such as parenthesis, commas and brackets, are no longer necessary, as the structure of the tree conveys this semantic information.

@@

\section{Values and expressions}
The main purpose of most computer software is to handle and manipulate data.
Each of these individual pieces of data is a value, which could be anything from the number 4 or the text `Hello!'.

@@

\subsubsection{RAG Modules}
\begin{figure}[h]
\begin{lstlisting}[language=jrag]
aspect TypingRules {
  syn Type Zero.type() {
    return new Int();
  }
}
\end{lstlisting}
  \caption{Snippet of a .jrag file, declaring a type attribute for the Zero AST node}
  \label{boolstrgen}
\end{figure}
The RAG modules written in \lstinline{.jrag} files, declare the attributes to be included in the classes generated by the AST specification.
The files use a syntax very similar to that of Java to declare attributes of a variety of different kinds, and the body of methods corresponds directly to Java code.

@@

uppercase_id = "[A-Z][A-Za-z0-9-]";
uppercase_id = "[a-z][A-Za-z0-9-]";
\end{lstlisting}
  \caption{A definition of the concrete syntax for the typing rule language in a notation similar to BNF.}
  \label{trastspec}
\end{figure}
%\CR{After reading, I would now say that it's preferable to me if you keep the language definition separate
%  from its implementation.  You can use Figure 3.4 as the abstract syntax, to complement the current discussion.
%  This would be before the current 3.1, then.
%  An advantage of this is that you can then start your ``Architecture'' discussion (in 3.2) by mentioning
%  the complex dependencies, i.e., that processing a type system specification requires you to know the abstract syntax
%  of the object language as well as of the types.}

\begin{figure}[h]
\centering
\begin{minipage}{.45\textwidth}
  \centering
  \begin{equation}
    \tag{T-True}
    \text{true} : \text{Bool}
  \end{equation}
  %\caption*{Pierce syntax}
  %\label{fig:sub1}
\end{minipage}%
\hspace{.1\textwidth}\begin{minipage}{.45\textwidth}
  \centering
\begin{lstlisting}[]
[T-True]
True : Bool
\end{lstlisting}
  %\caption*{ASCII representation}
  %\label{fig:sub2}
\end{minipage}
\caption{A typing rule for the value true in natural semantic syntax on the left and its equivalent ASCII representation on the right}
\label{trexampletrue}
\end{figure}
\begin{figure}[h]
\centering
\begin{minipage}{.45\textwidth}
  \centering
  \begin{equation}
    \tag{T-If}
    \frac{
      t_1 : \text{Bool}\quad
      t_2 : \text{T}\quad
      t_3 : \text{T}
    }{
      \text{if } t_1 \text{ then } t_2 \text{ else } t_3 : \text{T}
    }
  \end{equation}
  %\caption*{Pierce syntax}
  %\label{fig:sub1}
\end{minipage}%
\hspace{.1\textwidth}\begin{minipage}{.45\textwidth}
  \centering
\begin{lstlisting}[]
[T-If]
t1 : Bool,
t2 : t,
t3 : t
-----------------
If(t1, t2, t3): t
\end{lstlisting}
  %\caption*{ASCII representation}
  %\label{fig:sub2}
\end{minipage}
\caption{A typing rule describing an if-expression in natural semantic syntax and its equivalent ASCII representation}
\label{trexampleif}
\end{figure}

@@

In this section we will present the code we generate to implement the previously discussed algorithm.
Once typing rules have been parsed, we end up with an abstract syntax tree, in the form of Figure \ref{trastspec}, where the \lstinline{RuleSet} is the root node.
From the root node, calls propagate down the tree to fill in increasingly specific sections of the code.
%TODO: Rewrite this section to be less bad

This section details the implementation of our type checker generator, on a feature by feature basis.

@@

In Section \ref{sectionresearchquestions}, we evaluate the overall result of our study by answering our research questions.
%\CR{The section has two parts, so the intro paragraph here can directly discuss their purpose.}
%\section{Experimental Setup} % should describe the details of the method used to evaluate your solution(s)/approach. Sometimes this is already addressed in the \textbf{Method}, sometimes this part replaces \textbf{Method}.
\section{Results} % contains the data (as tables, graphs) obtained via experiments  (benchmarking, polls, interviews). Here you should also describe the individual tables or graphs in text, pointing out interesting outliers and trends.
We have chosen to evaluate our program using two example languages, inspired by the languages Booleans and Numbers introduced in Types and Programming Languages\ref{Pierce}.
We have chosen to simply the languages somewhat, to reduce the amount of equivalent cases for our program to parse, making it easier to evaluate each feature individually.
Thus, we have defined two languages Bools and BoolsIf, where the former is a simple language consisting of boolean values, an \lstinline{OR} operator, and a zero-value.
This is a minimal case to produce a meaningful type checker, as it contains values of two different types and a rule with premises limiting the types.

@@

%\begin{figure}
%\centering
%  \begin{minipage}[b]{.4\textwidth}
%  \centering
%\begin{lstlisting}[]
%[T-If]
%x: Bool,
%y: a,
%z: a
%---------
%If(x, y, z): a
%\end{lstlisting}
%  \captionof{figure}{A figure}
%  \label{fig:test1}
%\end{minipage}%
%\begin{minipage}[b]{.6\textwidth}
%  \centering
%\begin{lstlisting}[]
%syn Type If.type() {
%ASTNode x = getChild(0);
%ASTNode y = getChild(1);
%ASTNode z = getChild(2);
%
%Type tyvar_a = null;
%if(tyvar_a == null) tyvar_a = y.type();
%else if (!tyvar_a.matches(y.type()))
%throw new RuntimeException("Typechecking failed: Type variable mismatch");
%if(tyvar_a == null) tyvar_a = z.type();
%else if (!tyvar_a.matches(z.type()))
%throw new RuntimeException("Typechecking failed: Type variable mismatch");
%if(x.type().matches(new Bool()) && y.type().matches(tyvar_a) && z.type().matches(tyvar_a)) {
%return tyvar_a;
%}
%throw new RuntimeException("Typechecking failed");
%}
%\end{lstlisting}
%  \captionof{figure}{Another figure}
%  \label{fig:test2}
%\end{minipage}
%\end{figure}
As an example of rules utilising type variables, we have extended the Bools language with an if term, as used in Pierce's Booleans language\ref{Pierce}.
This if term captures each of the utilities of type variables in typing rules, using it both to ensure that multiple terms are of the same type and then using that same type as the result of the rule evaluation.
%\CR{``Can extend'' sounds a bit arbitrary.  Can you motivate with Pierce (even if not exactly)?}
As shown in the AST extension in Figure \ref{ifast}, it consists of three terms: a conditional, a term to be evaluated if the condition is true and a condition to evaluate if false.
A typing rule of this if term can be seen in Figure \ref{iftr}.
The rule has three premises, the first of which declares the conditional must have the type \lstinline{Bool}.
The next two declare that the two other children must be of the type \lstinline{a}, with the lowercase identifier indicating a type variable.
This type variable \lstinline{a} also subsequently appears in the resulting type of the if term.

@@

A cleaner solution to this might be view rules as a kind of pattern matching, where each rule is evaluated one by one until a matching one is found.
Instead of reporting the reason any individual rule failed, the error might instead display either all or only the best matching of the rules it tried to apply, perhaps with a list of non-holding premises for each one.
%\CR{This is a good idea; you can even sort the rules by closest matches and only show the top ones if there are many.}

@@

\subsubsection{Error handling}
We chose a very simple representation for type errors, based around runtime exceptions, which are thrown either when type variable unification fails or the premises of a typing rule fails to match.
The exceptions include a short message stating which of the two causes has lead to the error.

This approach gave us the minimum information to distinguish between successful and failing type checking, with a very simple implementation.
It does however complicate the case where a single node has multiple applicable typing rules, depending on their premises.
In Section \ref{improvederrorhandling} we discuss alternative approaches.

@@

\begin{appendices} %should contain lengthy details of the experimental setup, mathematical proofs, code download information, and shorter code snippets. Avoid longer code listings. Source code should rather be made available for download on a website or on-line repository of your choosing.
\chapter{Source code}
\url{https://github.com/nichobi/thesis-project}

@@

%make sure we're on even page with the pop-sci
\checkoddpage
\ifoddpage
\else
  \newpage
  \thispagestyle{empty}
  \mbox{ }
\fi
\includepdf[pages={1}]{popsci/popsci.pdf}
\end{appendices}

>>>>

  This report contains a study on the usability of JastAdd to programmatically generate executable type checkers from a subset of possible typing rules, known as 'syntax-directed' typing rules.
  This aims to provide integration with existing compilers implemented with JastAdd, as well as a playground for type systems usable as an educational tool.

  % 'Syntax-directed' needs explaining if used, "a subset known as syntax-directed"
  We find RAGs to be a flexible model for syntax-directed typing rules, and our software was able to support small languages from research literature.
  It proved difficult to provide compile-time verification of the typing rules, and further work is required to extend the compiler to support type environments, a necessity for most real-world type systems.
}

@@

\section{Abstract syntax trees}\label{abstractsyntaxtrees}
The syntax of a language is the definition of the symbols and structures which make up a valid program.
%\CR{What's a syntax?}
An abstract syntax tree (AST) is a simplified tree-based representation of the syntax.
Each node in the tree represents a syntactic construct of the language, such as expressions, statements or declarations and the branches represent the structure.
It is abstract in that it neglects to include all the concrete details of the syntax, but rather focuses on the structure of the program.
Structural elements of the source code, such as parenthesis, commas and brackets, are no longer necessary, as the structure of the tree conveys this semantic information.

@@

\subsection{Values and expressions}
The main purpose of most computer software is to handle and manipulate data.
Each of these individual pieces of data is a value, which could be anything from the number 4 or the text `Hello!'.

@@

\subsubsection{RAG Modules}
\begin{figure}[h]
\begin{lstlisting}[language=jrag]
aspect TypingRules {
  syn Type Zero.type() {
    return new Int();
  }
}
\end{lstlisting}
  \caption{Snippet of a .jrag file, declaring a type attribute for the Zero AST node}
  \label{jragexample}
\end{figure}
The RAG modules are written in \lstinline{.jrag} files, declare the attributes to be included in the classes generated by the AST specification.
Figure \ref{jragexample} shows a minimal example.
The files use a syntax very similar to that of Java to declare attributes of a variety of different kinds, and the body of methods corresponds directly to Java code.

@@

uppercase_id = "[A-Z][A-Za-z0-9-]";
uppercase_id = "[a-z][A-Za-z0-9-]";
\end{lstlisting}
  \caption{A definition of the concrete syntax for the typing rule language in a notation similar to BNF.}
  \label{trconcretesyntaxspec}
\end{figure}
%\CR{After reading, I would now say that it's preferable to me if you keep the language definition separate
%  from its implementation.  You can use Figure 3.4 as the abstract syntax, to complement the current discussion.
%  This would be before the current 3.1, then.
%  An advantage of this is that you can then start your ``Architecture'' discussion (in 3.2) by mentioning
%  the complex dependencies, i.e., that processing a type system specification requires you to know the abstract syntax
%  of the object language as well as of the types.}

\begin{figure}[]
\centering
\begin{minipage}{.45\textwidth}
  \centering
  \begin{equation}
    \tag{T-True}
    \text{true} : \text{Bool}
  \end{equation}
  %\caption*{Pierce syntax}
  %\label{fig:sub1}
\end{minipage}%
\hspace{.1\textwidth}\begin{minipage}{.45\textwidth}
  \centering
\begin{lstlisting}[]
[T-True]
True : Bool
\end{lstlisting}
  %\caption*{ASCII representation}
  %\label{fig:sub2}
\end{minipage}
\caption{A typing rule for the value true in natural semantic syntax on the left and its equivalent ASCII representation on the right}
\label{trexampletrue}
\end{figure}
\begin{figure}[]
\centering
\begin{minipage}{.45\textwidth}
  \centering
  \begin{equation}
    \tag{T-If}
    \frac{
      t_1 : \text{Bool}\quad
      t_2 : \text{T}\quad
      t_3 : \text{T}
    }{
      \text{if } t_1 \text{ then } t_2 \text{ else } t_3 : \text{T}
    }
  \end{equation}
  %\caption*{Pierce syntax}
  %\label{fig:sub1}
\end{minipage}%
\hspace{.1\textwidth}\begin{minipage}{.45\textwidth}
  \centering
\begin{lstlisting}[]
[T-If]
t1 : Bool,
t2 : t,
t3 : t
-----------------
If(t1, t2, t3): t
\end{lstlisting}
  %\caption*{ASCII representation}
  %\label{fig:sub2}
\end{minipage}
\caption{A typing rule describing an if-expression in natural semantic syntax and its equivalent ASCII representation}
\label{trexampleif}
\end{figure}

@@

In this section we will present the code we generate to implement the previously discussed algorithm.
Once typing rules have been parsed, we end up with an abstract syntax tree, in the form of Figure \ref{trastspec}, where the \lstinline{RuleSet} is the root node.
From the root node, calls propagate down the tree to fill in increasingly specific sections of the code.

%This section details the implementation of our type checker generator, on a feature by feature basis.

@@

In Section \ref{sectionresearchquestions}, we evaluate the overall result of our study by answering our research questions.
%\CR{The section has two parts, so the intro paragraph here can directly discuss their purpose.}
%\section{Experimental Setup} % should describe the details of the method used to evaluate your solution(s)/approach. Sometimes this is already addressed in the \textbf{Method}, sometimes this part replaces \textbf{Method}.
\section{Results} % contains the data (as tables, graphs) obtained via experiments  (benchmarking, polls, interviews). Here you should also describe the individual tables or graphs in text, pointing out interesting outliers and trends.
We have chosen to evaluate our program using two example languages, inspired by the languages Booleans and Numbers introduced in Types and Programming Languages\cite{Pierce}.
We have chosen to simply the languages somewhat, to reduce the amount of equivalent cases for our program to parse, making it easier to evaluate each feature individually.
Thus, we have defined two languages Bools and BoolsIf, where the former is a simple language consisting of boolean values, an \lstinline{OR} operator, and a zero-value.
This is a minimal case to produce a meaningful type checker, as it contains values of two different types and a rule with premises limiting the types.

@@

%\begin{figure}
%\centering
%  \begin{minipage}[b]{.4\textwidth}
%  \centering
%\begin{lstlisting}[]
%[T-If]
%x: Bool,
%y: a,
%z: a
%---------
%If(x, y, z): a
%\end{lstlisting}
%  \captionof{figure}{A figure}
%  \label{fig:test1}
%\end{minipage}%
%\begin{minipage}[b]{.6\textwidth}
%  \centering
%\begin{lstlisting}[]
%syn Type If.type() {
%ASTNode x = getChild(0);
%ASTNode y = getChild(1);
%ASTNode z = getChild(2);
%
%Type tyvar_a = null;
%if(tyvar_a == null) tyvar_a = y.type();
%else if (!tyvar_a.matches(y.type()))
%throw new RuntimeException("Typechecking failed: Type variable mismatch");
%if(tyvar_a == null) tyvar_a = z.type();
%else if (!tyvar_a.matches(z.type()))
%throw new RuntimeException("Typechecking failed: Type variable mismatch");
%if(x.type().matches(new Bool()) && y.type().matches(tyvar_a) && z.type().matches(tyvar_a)) {
%return tyvar_a;
%}
%throw new RuntimeException("Typechecking failed");
%}
%\end{lstlisting}
%  \captionof{figure}{Another figure}
%  \label{fig:test2}
%\end{minipage}
%\end{figure}
As an example of rules utilising type variables, we have extended the Bools language with an if term, as used in Pierce's Booleans language\cite{Pierce}.
This if term captures each of the utilities of type variables in typing rules, using it both to ensure that multiple terms are of the same type and then using that same type as the result of the rule evaluation.
%\CR{``Can extend'' sounds a bit arbitrary.  Can you motivate with Pierce (even if not exactly)?}
As shown in the AST extension in Figure \ref{ifast}, it consists of three terms: a conditional, a term to be evaluated if the condition is true and a condition to evaluate if false.
A typing rule of this if term can be seen in Figure \ref{iftr}.
The rule has three premises, the first of which declares the conditional must have the type \lstinline{Bool}.
The next two declare that the two other children must be of the type \lstinline{a}, with the lowercase identifier indicating a type variable.
This type variable \lstinline{a} also subsequently appears in the resulting type of the if term.

@@

A more informative solution to this might be view rules as a kind of pattern matching, where each rule is evaluated one by one until a matching one is found.
Instead of reporting the reason any individual rule failed, the error might instead display either all or only the best matching of the rules it tried to apply, perhaps with a list of non-holding premises for each one.
%\CR{This is a good idea; you can even sort the rules by closest matches and only show the top ones if there are many.}
%\subsubsection{Error handling}
%We chose a very simple representation for type errors, based around runtime exceptions, which are thrown either when type variable unification fails or the premises of a typing rule fails to match.
%The exceptions include a short message stating which of the two causes has lead to the error.
%
%This approach gave us the minimum information to distinguish between successful and failing type checking, with a very simple implementation.
%It does however complicate the case where a single node has multiple applicable typing rules, depending on their premises.
%In Section \ref{improvederrorhandling} we discuss alternative approaches.

@@

@@

\begin{appendices} %should contain lengthy details of the experimental setup, mathematical proofs, code download information, and shorter code snippets. Avoid longer code listings. Source code should rather be made available for download on a website or on-line repository of your choosing.
\chapter{Source code}
The source code for our project is available at \\\url{https://github.com/nichobi/thesis-project}, \\with the README.md file providing instructions for use.

\vspace*{\fill}
\hspace*{\fill}
\includegraphics[scale=0.2]{stagbeetle.pdf}

@@

%TODO: Re-add the popsci summary
%%make sure we're on even page with the pop-sci
%\checkoddpage
%\ifoddpage
%\else
%  \newpage
%  \thispagestyle{empty}
%  \mbox{ }
%\fi
%\includepdf[pages={1}]{popsci/popsci.pdf}
\end{appendices}
